Estimating species interactions in large, abiotically structured communities with Markov networks and stochastic approximation

### 2016-02-11 stochastic/stochastic-recovery.Rmd
some chapter 3 work

```{r,eval=FALSE}
library(rosalia)
library(mistnet)

@@

```{r,eval=FALSE}
set.seed(1)
n_spp = 250     # number of species
n_loc = 2500    # number of locations
n_env = 5       # number of environmental predictors
n_gibbs = 5000  # number of Gibbs sampling iterations

@@

```{r,eval=FALSE, cache = TRUE}
# Initialize the `y` matrix.
# This will hold the "observed" presence-absence data
y = matrix(0.5, nrow = n_loc, ncol = n_spp)

@@

```{r,eval=FALSE}
# Calculate sufficient statistics of the data
y_stats = crossprod(y)
y_env_stats = t(true_env) %*% y
```

```{r,eval=FALSE}
# Initialize the simulated landscape for stochastic approximation
y_sim = matrix(0.5, nrow = nrow(y), ncol = ncol(y))

@@

```{r,eval=FALSE}
# Very weak priors on alpha terms, somewhat stronger on beta terms
alpha_env_prior = rosalia::make_logistic_prior(scale = 2)$log_grad 
alpha_species_prior = rosalia::make_logistic_prior(scale = 2)$log_grad
beta_prior = rosalia::make_logistic_prior(scale = 0.5)$log_grad
```

```{r,eval=FALSE}
initial_learning_rate = 1 # step size at start of optimization
maxit = 50000             # Number of rounds of optimization
start_time = as.integer(Sys.time())

@@

```{r,eval=FALSE}
library(cowplot)
library(ggplot2)
out = plot_grid(
  ggplot(NULL, aes(x = beta[upper.tri(beta)], 
                   y = true_beta[upper.tri(beta)])) + 
    stat_binhex() + 
    xlab("Estimated coefficient value") + 
    ylab("\"True\" coefficient value") + 
    stat_hline(yintercept = 0, size = 1/8) + 
    stat_vline(xintercept = 0, size = 1/8) + 
    scale_fill_gradient(low = "#F0F0F0", high = "darkblue", 
                        trans = "log10") +
    stat_abline(intercept = 0, slope = 1, size = 1/2) + 
    coord_equal(),  
  ggplot(NULL, aes(x = c(alpha_env), y = c(true_alpha_env))) + 
    stat_binhex() + 
    xlab("Estimated coefficient value") + 
    ylab("\"True\" coefficient value") + 
    stat_hline(yintercept = 0, size = 1/8) + 
    stat_vline(xintercept = 0, size = 1/8) + 
    scale_fill_gradient(low = "#F0F0F0", high = "darkblue", 
                        trans = "log10") +
    stat_abline(intercept = 0, slope = 1, size = 1/2) + 
    coord_equal(),  
  nrow = 1,
  labels = c(
    "A. Pairwise biotic coefficients (n = 31125)", 
    "B. Abiotic coefficients (n = 1250)"
  )
)
save_plot(
  "stochastic/estimates.pdf", 
  out, 
  base_aspect_ratio = 2, 
  base_height = 6
)

@@

```{r,eval=FALSE}
small_n_spp = 20
small_n_loc = 500
small_true_beta = true_beta[1:small_n_spp, 1:small_n_spp]
small_true_alpha = true_alpha_species - 2
small_y = matrix(0.5, nrow = small_n_loc, ncol = small_n_spp)

@@

```{r,eval=FALSE}
exact_time = system.time({
  exact = rosalia(small_y, 
                  prior = make_logistic_prior(scale = 1), maxit = 500)
})

@@

```{r,eval=FALSE}
maxit_small = 50000
start_time_small = as.numeric(Sys.time())

@@

```{r,eval=FALSE}
pdf("stochastic/convergence.pdf", width = 10, height = 7)
par(mfrow = c(1, 2), las = 1)

plot(small_times, mses, type = "l", 
     ylab = "mean square deviation from MLE\n(log scale)", 
     xlab = "time (seconds)", 
     main = "A. Convergence to known MLE\n(small network)", 
     yaxs = "i", bty = "l", log = "y", xaxs = "i", lwd = 2, 
     col = "blueviolet", xlim = range(c(-.1, small_times)))

>>>>

```{r,cache = TRUE}
library(rosalia)
library(mistnet)

@@

```{r,cache = TRUE}
set.seed(1)
n_spp = 250     # number of species
n_loc = 2500    # number of locations
n_env = 5       # number of environmental predictors
n_gibbs = 5000  # number of Gibbs sampling iterations

@@

```{r,, cache = TRUE}
# Initialize the `y` matrix.
# This will hold the "observed" presence-absence data
y = matrix(0.5, nrow = n_loc, ncol = n_spp)

@@

```{r,cache = TRUE}
# Calculate sufficient statistics of the data
y_stats = crossprod(y)
y_env_stats = t(true_env) %*% y
```

```{r,cache = TRUE}
# Initialize the simulated landscape for stochastic approximation
y_sim = matrix(0.5, nrow = nrow(y), ncol = ncol(y))

@@

```{r,cache = TRUE}
# Very weak priors on alpha terms, somewhat stronger on beta terms
alpha_env_prior = rosalia::make_logistic_prior(scale = 2)$log_grad 
alpha_species_prior = rosalia::make_logistic_prior(scale = 2)$log_grad
beta_prior = rosalia::make_logistic_prior(scale = 0.5)$log_grad
```

```{r,cache = TRUE}
initial_learning_rate = 1 # step size at start of optimization
maxit = 50000             # Number of rounds of optimization
start_time = as.integer(Sys.time())

@@

```{r,cache = TRUE}
library(cowplot)
library(ggplot2)
out = plot_grid(
  ggplot(NULL, aes(x = beta[upper.tri(beta)], 
                   y = true_beta[upper.tri(beta)])) + 
    stat_binhex() + 
    xlab("Estimated coefficient value") + 
    ylab("\"True\" coefficient value") + 
    stat_hline(yintercept = 0, size = 1/8) + 
    stat_vline(xintercept = 0, size = 1/8) + 
    scale_fill_gradient(low = "#F0F0F0", high = "darkblue", 
                        trans = "log10") +
    stat_abline(intercept = 0, slope = 1, size = 1/2) + 
    coord_equal(),

  ggplot(NULL, aes(x = c(alpha_env), y = c(true_alpha_env))) + 
    stat_binhex() + 
    xlab("Estimated coefficient value") + 
    ylab("\"True\" coefficient value") + 
    stat_hline(yintercept = 0, size = 1/8) + 
    stat_vline(xintercept = 0, size = 1/8) + 
    scale_fill_gradient(low = "#F0F0F0", high = "darkblue", 
                        trans = "log10") +
    stat_abline(intercept = 0, slope = 1, size = 1/2) + 
    coord_equal(),  
  nrow = 1,
  labels = c(
    "A. Pairwise biotic coefficients (n = 31125)", 
    "B. Abiotic coefficients (n = 1250)"
  )
)
save_plot(
  "stochastic/estimates.pdf", 
  out, 
  base_aspect_ratio = 2, 
  base_height = 6
)

@@

```{r,cache = TRUE}
small_n_spp = 20
small_n_loc = 500
small_true_beta = true_beta[1:small_n_spp, 1:small_n_spp]
small_true_alpha = true_alpha_species - 2
small_y = matrix(0.5, nrow = small_n_loc, ncol = small_n_spp)

@@

```{r,cache = TRUE}
exact_time = system.time({
  exact = rosalia(small_y, 
                  prior = make_logistic_prior(scale = 1), maxit = 500)
})

@@

```{r,cache = TRUE}
maxit_small = 50000
start_time_small = as.numeric(Sys.time())

@@

```{r,cache = TRUE}
pdf("stochastic/convergence.pdf", width = 10, height = 7)
par(mfrow = c(1, 2), las = 1)

plot(small_times, mses, type = "l", 
     ylab = "mean square deviation from MLE", 
     xlab = "time (seconds)", 
     main = "A. Convergence to known MLE\n(small network)", 
     yaxs = "i", bty = "l", log = "y", xaxs = "i", lwd = 2, 
     col = "blueviolet", xlim = range(c(-.1, small_times)))
